Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-02-13 newpaper/paper.tex
stuff

\newcommand{\Int}{\msf{Int}}
\newcommand{\List}{\msf{List}\,}
\newcommand{\Nil}{\msf{Nil}}
\newcommand{\Cons}{\msf{Cons}}
\newcommand{\Reader}{\msf{Reader}}
\newcommand{\ReaderT}{\msf{ReaderT}}
\newcommand{\Monad}{\msf{Monad}}
\newcommand{\Applicative}{\msf{Monad}}
\newcommand{\class}{\msf{class}}
\newcommand{\Functor}{\msf{Functor}}
\newcommand{\Bool}{\msf{Bool}}
\newcommand{\Statel}{\msf{State}}
\newcommand{\fro}{\leftarrow}
\newcommand{\case}{\mbf{case\,}}

@@

\theoremstyle{remark}
\newtheorem{notation}{Notation}
\newcommand{\id}{\mit{id}}
\newcommand{\mup}{\mit{up}}
\newcommand{\mdown}{\mit{down}}
\newcommand{\tyclass}{\mbf{class}}
\newcommand{\instance}{\mbf{instance}\,}
\newcommand{\Improve}{\msf{Improve}}
\newcommand{\Gen}{\msf{Gen}}
\newcommand{\unGen}{\mit{unGen}}
\renewcommand{\Vec}{\msf{Vec}}
\newcommand{\gen}{\mit{gen}}
\newcommand{\genRec}{\mit{genRec}}
\newcommand{\fmap}{<\!\!\$\!\!>}
\newcommand{\ap}{<\!\!*\!\!>}
\newcommand{\runGen}{\mit{runGen}}
\newcommand{\qt}[1]{\ql#1\qr}
\newcommand{\lift}{\mit{lift}}
\newcommand{\liftGen}{\mit{liftGen}}
\newcommand{\MonadGen}{\msf{MonadGen}}
\newcommand{\RA}{\Rightarrow}
\newcommand{\EitherT}{\msf{EitherT}}
\newcommand{\StateT}{\msf{StateT}}
\newcommand{\Identity}{\msf{Identity}}
\newcommand{\runIdentity}{\mit{runIdentity}}
\newcommand{\newtype}{\mbf{newtype}\,}
\newcommand{\runMaybeT}{\mit{runMaybeT}}
\newcommand{\dlr}{\,\$\,}
\newcommand{\ImproveF}{\msf{ImproveF}}
\newcommand{\ExceptT}{\msf{ExceptT}}

@@

@@

@@

Inductive types can be introduced using a Haskell-like ADT notation, or with a GADT-style one:
\begin{alignat*}{3}
  &                                           &&\hspace{4em}\data\,\Bool_\M &&: \MTy\,\where\\
  & \data\,\Bool_\M : \MTy = \True\,|\,\False &&\hspace{4em}\ind\ind \True &&: \Bool_\M\\
  &                                           &&\hspace{4em}\ind\ind \False &&: \Bool_\M
\end{alignat*}
Note that we added an $_\M$ subscript to the type; when analogous types can be
defined both on the meta and object levels, we will sometimes use this subscript
to disambiguate the meta-level version. We will also use Haskell-like newtype
notation, such as in $\newtype \msf{Wrap}\,A = \msf{Wrap}\,\{\mit{unWrap} : A
\}$.

@@

Object-level definitions use $:=$ as notation, instead of the $=$ that is used
for meta-level ones. Later on, we may use some implicit CFTT notations, but we
will always disambiguate the stages of $\mbf{let}$-s in this way. Non-recursive
$\mbf{let}$ is also allowed, and can be used to shadow binders:
\begin{alignat*}{3}
  &f : \Nat \to \Nat\\
  &f\,x := \letdef x := x + 10; \letdef x := x + 20; x * 10
\end{alignat*}
Values are call-by-value at runtime; they are computed eagerly in function
applications and $\mbf{let}$-s. $\mbf{let}$-definitions can be used to define
inhabitants of any type, and the type of the $\mbf{let}$ body can be also
arbitrary. Additionally, the right hand sides of $\case$ branches can also have
arbitrary types. So the following is well-formed:
\begin{alignat*}{3}
  &\hspace{-1em}f : \mathrlap{\Bool \to \Nat \to \Nat}\\
  &\hspace{-1em}f\,b\,\mathrlap{:= \case b\,\of}\\
  &\hspace{-1em}\ind \True  &&\to \lam x.\, x + 10;\\
  &\hspace{-1em}\ind \False &&\to \lam x.\, x * 10;
\end{alignat*}

@@

Let us look at some basic staging examples. Recall the meta-level identity
function; it can be used at the object-level too, by applying it to quoted
terms:
  \[ \letdef n : \Nat := \spl(\id\,\ql 10 + 10 \qr); ... \]
Here, $\id$ is used at type $\Lift \Nat$. During staging, the expression in the
splice is evaluated, so we get $\spl\ql 10 + 10 \qr$, which is definitionally
the same as $10 + 10$, which is our staging output here. Boolean
short-circuiting is another basic use-case:
\begin{alignat*}{3}
  &\mit{and} : \Up\Bool \to \Up\Bool \to \Up\Bool\\
  &\mit{and}\,x\,y = \ql\case \spl x\,\of \True \to \spl y; \False \to \False\qr
\end{alignat*}
Since the $y$ expression is inlined under a $\case$ branch at every use site, it
is only computed at runtime when $x$ evaluates to $\True$. In many situations,
staging can be used instead of laziness to implement short-circuiting, and with
generally better runtime performance, avoiding the overhead of thunking. Consider
the $\map$ function now:
\begin{alignat*}{3}
  & \hspace{-5em}\mathrlap{\map : \{A\,B : \VTy\} \to (\Up A \to \Up B) \to \Up (\List A) \to \Up(\List B)}\\
  & \hspace{-5em}\mathrlap{\map\,f\,\vas = \ql\letrec \go\,\vas := \case \vas\,\of}\\
  & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Nil           &&\to \Nil;\\
  & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Cons\,a\,\vas &&\to \Cons\,\spl(f\,\ql a \qr)\,(\go\,\vas);\\
  & \hspace{-5em}\ind\ind\ind\ind\ind\go\,\spl\vas \qr
\end{alignat*}
For example, this can be used as $\letdef f\,\vas : \List \Nat \to \List \Nat :=
\spl(\map\,(\lam x.\,\ql \spl x + 10 \qr)\,\qt{\vas}) $.  This is staged to a recursive
definition where the mapping function is inlined into the $\Cons$ case as
$\Cons\,a\,\vas \to \Cons\,(a + 10)\,(\go\,\vas)$.  Note that $\map$ has to
abstract over value types, since lists can only contain values, not functions. Also, the
mapping function has type $\Up A \to \Up B$, instead of $\Up (A \to B)$. The
former type is often preferable to the latter in staging; the former is
a metafunction with useful computational content, while the latter is merely a
black box that computes object code. If we have $f : \Up(A \to B)$, and $f$ is
staged to $\ql \lam x.\,t\qr$, then $\spl f u$ is staged to an
undesirable ``administrative'' $\beta$-redex $(\lam x.\,t)\,u$.

\section{Staged Programming in CFTT}\label{sec:abstractions-in-cftt}

@@

We saw that $\Up A \to \Up B$ is usually preferable to $\Up(A \to B)$. The two
types are actually equivalent up to the runtime semantics of object programs,
and we can convert back and forth in CFTT:
\begin{alignat*}{3}
  &\mup : \Up (A \to B) \to \Up A \to \Up B && \ind\ind \mdown : (\Up A \to \Up B) \to \Up (A \to B) \\
  &\mup\,f\,a = \ql \spl f\, \spl a\qr   && \ind\ind \mdown\,f = \ql \lam a.\,\spl(f\,\ql a \qr) \qr
\end{alignat*}
We can not show internally, using propositional equality, that these functions are
inverses, since we do not have $\beta\eta$-rules for object functions; but we
will not need that in the rest of the paper.

@@

Our solution is to use a monad which freely extends $\MTy$ with the ability to
generate object-level code and introduce object binders.
\begin{alignat*}{3}
  & \newtype \Gen\,A = \Gen\,\{\unGen : \{R : \Ty\} \to (A \to \Up R) \to \Up R\}
\end{alignat*}
This is a monad in $\MTy$ in a very standard sense, and we reuse Haskell type classes
and $\mdo$-notation.
\begin{alignat*}{3}
  &\instance \Monad\,\Gen\,\where\\
  &\ind \return a = \Gen\,(\lam k.\,k\,a)\\
  &\ind \mit{ga} \bind f = \Gen\,(\lam k.\,\unGen\,\mit{ga}\,(\lam a.\,(f\,a\,k)))
\end{alignat*}
From the $\Monad$ instance, the $\Functor$ and $\Applicative$ instances can be also derived.
We will reuse $(\fmap)$ and $(\ap)$ for applicative notation
as well.
An inhabitant of $\Gen\,A$ can be viewed as an action whose effect is to produce
some object code and returns an $A$ value that can depend on produced object
binders. $\Gen$ can be only ``run'' when the result type is an object type:
\begin{alignat*}{3}
  &\runGen : \Gen\,(\Up A) \to \Up A\\
  &\runGen\,\vma = \unGen\,\vma\,\id
\end{alignat*}
We can let-bind object expressions in $\Gen$:
\begin{alignat*}{3}
  & \gen : \Up A \to \Gen\,(\Up A) \\
  & \gen\,a = \Gen\,(\lam k.\,\ql \letdef x : A := \spl a; \spl(k\,\ql x \qr) \qr)
\end{alignat*}
And also recursive function definitions:
\begin{alignat*}{3}
  & \genRec : \{A : \CTy\} \to (\Up A \to \Up A) \to \Gen\,(\Up A) \\
  & \genRec\,f = \Gen\,(\lam k.\,\qt{\letrec x : A := \spl(f\,\qt{x}); \spl(k\,\qt{x})})
\end{alignat*}
Now, using do-notation, we may write $\mdo \{x \fro \gen\,\qt{10 + 10}; y \fro
\gen\,\qt{20 + 20}\};\return \qt{x + y}\}$, for a $\Gen\,(\Up \Nat)$
action. Running this with $\runGen$ yields $\qt{\letdef x := 10 + 10; \letdef y
  := 20 + 20; x + y}$. We can also define a ``safer'' binding-time improvement for
products, using let-insertion:
\begin{alignat*}{3}
  &\hspace{-4em}\mathrlap{\instance \Improve\,(A,\,B)\,(\Gen\,(\Up A,\,\Up B))\,\where}\\
  &\hspace{-4em}\ind \mup\,x   && = \mdo\{x \fro \gen\,x; \return (\qt{\fst\,\spl x},\,\qt{\snd\,\spl x})\}\\
  &\hspace{-4em}\ind \mdown\,x && = \runGen\,\$\,\mdo \{(a,\,b) \fro x; \return \qt{(\spl a,\, \spl b)}\}
\end{alignat*}

Working in $\Gen$ is convenient, since we can freely generate object code and
also have access to the full metalanguage. Also, the whole point of staging is
that \emph{eventually} all metaprograms will be used for the purpose of code
generation, so ultimately we always want to $\runGen$ our actions. So why not
just always work in $\Gen$? We may not want to do that, because we would like to
reason about the size and content of the code we are generating, and the
implicit code generation in $\Gen$ is not very transparent. This is a bit
similar to the $\msf{IO}$ monad in Haskell, where eventually everything needs
to run in $\msf{IO}$, but we may prefer to not write most of our program in
$\msf{IO}$.

@@

Let us start with the $\Maybe$ monad. We have $\data \Maybe\,A : \VTy :=
\Nothing\,|\,\Just\,A$, and $\Maybe$ itself is available as a $\VTy \to \VTy$
metafunction. However, we cannot directly fashion a monad out of $\Maybe$, since
we do not have enough type formers in $\VTy$. We could try to use the following
type for binding:
\[ (\bind) : \Up(\Maybe\,A) \to (\Up\,A \to \Up\,(\Maybe\,B)) \to \Up(\Maybe\,B) \]
This works, but the definition necessarily uses runtime case splits on $\Maybe$
values, many of which could be optimized away during staging. Also, not having
a ``real'' monad is inconvenient for the purpose of code reuse.

Instead, our strategy is to only use ordinary monads in $\MTy$, and convert
between object types and meta-monads when necessary, as a form of binding-time
improvement. Assume that $\Maybe_M$ is a meta-level monad, and $\MaybeT_\M$ is
the monad transformer defined as follows:
\[ \newtype \MaybeT_\M\,M\,A = \MaybeT_\M\,\{\runMaybeT_\M : M\,(\Maybe_\M\,A)\} \]
Then the object-level $\Maybe$ is binding-time improved as follows:
\begin{alignat*}{3}
  &\instance \Improve\,(\Maybe\,A)\,(\MaybeT_\M\,\Gen\,(\Up\,A))\,\where\\
  %% &\ind \mup : \Up\,(\Maybe\,A) \to \MaybeT_\M\,\Gen\,(\Up\,A) \\
  &\ind \mup\,x = \MaybeT_\M \dlr \Gen \dlr \lam k.\\
  &\ind\ind \qt{\case \spl x\,\of \Nothing \to \spl(k\,\Nothing_\M);\Just\,a \to \spl(k\,(\Just_\M\qt{a}))}\\
  &\\
  %% &\ind \mdown : \MaybeT_\M\,\Gen\,(\Up\,A) \to \Up\,(\Maybe\,A) \\
  &\ind \mdown\,(\MaybeT_\M\,(\Gen\,\vma)) = \\
  &\ind\ind \vma\,(\lam x.\,\case x\,\of \Nothing_\M \to \qt{\Nothing}; \Just_\M\,a \to \qt{\Just\,\spl a})
\end{alignat*}
With this, we get the $\Monad$ instance for free from $\MaybeT_\M$ and $\Gen$. A small example:
\[ \letdef n : \Maybe\,\Nat := \spl(\mdown\,\$\,\mdo \{x \fro \return\,\qt{10}; y \fro \return\,\qt{20}; \return\,\qt{x + y})\});\,... \]
Since $\MaybeT_\M$ is meta-level, its binding always fully computes at staging time. Thus, the above code is staged to
\[ \letdef n : \Maybe\,\Nat := \Just\,(10 + 20);\;... \]
Assume also a $\lift : \Monad\,M \RA M\,A \to \MaybeT_\M\,M\,A$ operation, as in
Haskell. We shall assume this for all monad transformers henceforth. We can do
let-insertion in $\MaybeT_\M\,\Gen$ by simply lifting:
\begin{alignat*}{3}
  &\gen' : \Up A \to \MaybeT_\M\,\Gen\,(\Up A) \\
  &\gen'\,a = \lift\,(\gen\,a)
\end{alignat*}
However, it is more convenient to proceed in the style of Haskell's monad
transformer library \cite{TODO}, and have a class for monads that can do code
generation:
\begin{alignat*}{3}
&\tyclass\,\Monad\,M \RA \MonadGen\,M\,\where\\
&\ind \liftGen : \Gen\,A \to \M\,A\\
&\instance \MonadGen\,\Gen\,\where\,\liftGen = \id \\
&\instance \MonadGen\,M \RA \MonadGen\,(\MaybeT_\M\,M)\,\where\,\liftGen = \lift \circ \liftGen
\end{alignat*}
We also redefine $\gen$ and $\genRec$ to work in any $\MonadGen$, so from now on we have:
\begin{alignat*}{3}
 & \gen   &&: \MonadGen\,M \RA \Up A \to M\,(\Up A)\\
 &\genRec &&: \MonadGen\,M \RA (\Up A \to \Up A) \to M\,(\Up A)
\end{alignat*}

>>>>

\newcommand{\Int}{\msf{Int}}
\newcommand{\List}{\msf{List}}
\newcommand{\Nil}{\msf{Nil}}
\newcommand{\Cons}{\msf{Cons}}
\newcommand{\Reader}{\msf{Reader}}
\newcommand{\ReaderT}{\msf{ReaderT}}
\newcommand{\Monad}{\msf{Monad}}
\newcommand{\Applicative}{\msf{Monad}}
\newcommand{\class}{\msf{class}}
\newcommand{\Functor}{\msf{Functor}}
\newcommand{\Bool}{\msf{Bool}}
\newcommand{\Statel}{\msf{State}}
\newcommand{\fro}{\leftarrow}
\newcommand{\case}{\mbf{case\,}}

@@

\theoremstyle{remark}
\newtheorem{notation}{Notation}
\newcommand{\id}{\mit{id}}
\newcommand{\mup}{\mit{up}}
\newcommand{\mdown}{\mit{down}}
\newcommand{\tyclass}{\mbf{class}}
\newcommand{\instance}{\mbf{instance}\,}
\newcommand{\Improve}{\msf{Improve}}
\newcommand{\Gen}{\msf{Gen}}
\newcommand{\unGen}{\mit{unGen}}
\renewcommand{\Vec}{\msf{Vec}}
\newcommand{\gen}{\mit{gen}}
\newcommand{\genRec}{\mit{genRec}}
\newcommand{\fmap}{<\!\!\$\!\!>}
\newcommand{\ap}{<\!\!*\!\!>}
\newcommand{\runGen}{\mit{runGen}}
\newcommand{\qt}[1]{\ql#1\qr}
\newcommand{\lift}{\mit{lift}}
\newcommand{\liftGen}{\mit{liftGen}}
\newcommand{\MonadGen}{\msf{MonadGen}}
\newcommand{\MonadState}{\msf{MonadState}}
\newcommand{\MonadReader}{\msf{MonadReader}}
\newcommand{\RA}{\Rightarrow}
\newcommand{\EitherT}{\msf{EitherT}}
\newcommand{\StateT}{\msf{StateT}}
\newcommand{\Identity}{\msf{Identity}}

\newcommand{\Stop}{\msf{Stop}}
\newcommand{\Skip}{\msf{Skip}}
\newcommand{\Yield}{\msf{Yield}}

\newcommand{\runIdentity}{\mit{runIdentity}}
\newcommand{\runReaderT}{\mit{runReaderT}}
\newcommand{\newtype}{\mbf{newtype}\,}
\newcommand{\runMaybeT}{\mit{runMaybeT}}
\newcommand{\runStateT}{\mit{runStateT}}
\newcommand{\runState}{\mit{runState}}
\newcommand{\dlr}{\,\$\,}
\newcommand{\ImproveF}{\msf{ImproveF}}
\newcommand{\ExceptT}{\msf{ExceptT}}
\newcommand{\State}{\msf{State}}
\newcommand{\SumVS}{\msf{SumVS}}
\newcommand{\ProdCS}{\msf{ProdCS}}
\newcommand{\Here}{\msf{Here}}
\newcommand{\There}{\msf{There}}
\newcommand{\IsSumVS}{\msf{IsSumVS}}
\newcommand{\MonadJoin}{\msf{MonadJoin}}
\newcommand{\Stream}{\msf{Stream}}
\newcommand{\join}{\mit{join}}
\newcommand{\modify}{\mit{modify}}
\newcommand{\get}{\mit{get}}
\newcommand{\mput}{\mit{put}}
\newcommand{\mAs}{\mit{As}}
\newcommand{\Rep}{\mit{Rep}}
\newcommand{\encode}{\mit{encode}}
\newcommand{\decode}{\mit{decode}}
\newcommand{\mindex}{\mit{index}}
\newcommand{\mtabulate}{\mit{tabulate}}
\newcommand{\States}{\mit{States}}
\newcommand{\seed}{\mit{seed}}
\newcommand{\step}{\mit{step}}
\newcommand{\Step}{\msf{Step}}
\newcommand{\Pull}{\msf{Pull}}
\newcommand{\MkPull}{\msf{MkPull}}

@@

\todo{REORG intro, trim, contrib, motivate monadtrans \& stream fusion}

@@

\todo{If we won't use Sigma notation, get rid of it}

@@

Inductive types can be introduced using a Haskell-like ADT notation, or with a GADT-style one:
\begin{alignat*}{3}
  &                                           &&\hspace{4em}\data\,\Bool_\M &&: \MTy\,\where\\
  & \data\,\Bool_\M : \MTy = \True_\M\,|\,\False_\M &&\hspace{4em}\ind\ind \True_\M &&: \Bool_\M\\
  &                                           &&\hspace{4em}\ind\ind \False_\M &&: \Bool_\M
\end{alignat*}
Note that we added an $_\M$ subscript to the type; when analogous types can be
defined both on the meta and object levels, we will sometimes use this subscript
to disambiguate the meta-level version. We will also use Haskell-like newtype
notation, such as in $\newtype \msf{Wrap}\,A = \msf{Wrap}\,\{\mit{unWrap} : A
\}$.

@@

Object-level definitions use $:=$ as notation, instead of the $=$ that is used
for meta-level ones. Later on, we may use some implicit CFTT notations, but we
will always disambiguate the stages of $\mbf{let}$-s in this way. Non-recursive
$\mbf{let}$ is also allowed, and can be used to shadow binders:
\begin{alignat*}{3}
  &f : \Nat \to \Nat\\
  &f\,x := \letdef x := x + 10; \letdef x := x + 20; x * 10
\end{alignat*}
We also allow $\mbf{newtype}$ definitions, both in $\VTy$ and $\CTy$. These are
assumed to be erased at runtime. In the Haskell adaptation they are important
for aiding type class resolution, and we think that the explicit wrapping
makes many definitions more comprehensible in CFTT as well.

Values are call-by-value at runtime; they are computed eagerly in function
applications and $\mbf{let}$-s. $\mbf{let}$-definitions can be used to define
inhabitants of any type, and the type of the $\mbf{let}$ body can be also
arbitrary. Additionally, the right hand sides of $\case$ branches can also have
arbitrary types. So the following is well-formed:
\begin{alignat*}{3}
  &f : \mathrlap{\Bool \to \Nat \to \Nat}\\
  &f\,b := \case b\,\of \True \to (\lam x.\, x + 10); \False &&\to (\lam x.\, x * 10)
\end{alignat*}

@@

Let us look at some basic staging examples. Recall the meta-level identity
function; it can be used at the object-level too, by applying it to quoted
terms:
  \[ \letdef n : \Nat := \spl(\id\,\ql 10 + 10 \qr); ... \]
Here, $\id$ is used at type $\Lift \Nat$. During staging, the expression in the
splice is evaluated, so we get $\spl\ql 10 + 10 \qr$, which is definitionally
the same as $10 + 10$, which is our staging output here. Boolean
short-circuiting is another basic use-case:
\begin{alignat*}{3}
  &\mit{and} : \Up\Bool \to \Up\Bool \to \Up\Bool\\
  &\mit{and}\,x\,y = \ql\case \spl x\,\of \True \to \spl y; \False \to \False\qr
\end{alignat*}
Since the $y$ expression is inlined under a $\case$ branch at every use site, it
is only computed at runtime when $x$ evaluates to $\True$. In many situations,
staging can be used instead of laziness to implement short-circuiting, and with
generally better runtime performance, avoiding the overhead of thunking. Consider
the $\map$ function now:
\begin{alignat*}{3}
  & \hspace{-5em}\mathrlap{\map : \{A\,B : \VTy\} \to (\Up A \to \Up B) \to \Up (\List\,A) \to \Up(\List\,B)}\\
  & \hspace{-5em}\mathrlap{\map\,f\,\vas = \ql\letrec \go\,\vas := \case \vas\,\of}\\
  & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Nil           &&\to \Nil;\\
  & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Cons\,a\,\vas &&\to \Cons\,\spl(f\,\ql a \qr)\,(\go\,\vas);\\
  & \hspace{-5em}\ind\ind\ind\ind\ind\go\,\spl\vas \qr
\end{alignat*}
For example, this can be used as $\letdef f\,\vas : \List\,\Nat \to \List\,\Nat :=
\spl(\map\,(\lam x.\,\ql \spl x + 10 \qr)\,\qt{\vas}) $.  This is staged to a recursive
definition where the mapping function is inlined into the $\Cons$ case as
$\Cons\,a\,\vas \to \Cons\,(a + 10)\,(\go\,\vas)$.  Note that $\map$ has to
abstract over value types, since lists can only contain values, not functions. Also, the
mapping function has type $\Up A \to \Up B$, instead of $\Up (A \to B)$. The
former type is often preferable to the latter in staging; the former is
a metafunction with useful computational content, while the latter is merely a
black box that computes object code. If we have $f : \Up(A \to B)$, and $f$ is
staged to $\ql \lam x.\,t\qr$, then $\spl f u$ is staged to an
undesirable ``administrative'' $\beta$-redex $(\lam x.\,t)\,u$.

\section{Monads \& Monad Transformers}\label{sec:monad-transformers}

In this section we build a library for monads and monad transformers. We believe
that this is a good demonstration of CFTT-s abilities, since monads are
ubiquitous in practical Haskell programming, and they also introduce a great
amount of abstraction and that should be optimized away.

@@

We start with some preparatory work before getting to monads. We saw that $\Up
A \to \Up B$ is usually preferable to $\Up(A \to B)$. The two types are actually
equivalent up to the runtime semantics of object programs, and we can convert
back and forth in CFTT:
\begin{alignat*}{3}
  &\mup : \Up (A \to B) \to \Up A \to \Up B && \ind\ind \mdown : (\Up A \to \Up B) \to \Up (A \to B) \\
  &\mup\,f\,a = \ql \spl f\, \spl a\qr   && \ind\ind \mdown\,f = \ql \lam a.\,\spl(f\,\ql a \qr) \qr
\end{alignat*}
We can not show internally, using propositional equality, that these functions are
inverses, since we do not have $\beta\eta$-rules for object functions; but we
will not need this proof in the rest of the paper.

@@

Our solution is to use a monad which extends $\MTy$ with the ability to
freely generate object-level code and introduce object binders.
\begin{alignat*}{3}
  & \newtype \Gen\,A = \Gen\,\{\unGen : \{R : \Ty\} \to (A \to \Up R) \to \Up R\}
\end{alignat*}
This is a monad in $\MTy$ in a very standard sense, and we reuse Haskell type classes
and $\mdo$-notation.
\begin{alignat*}{3}
  &\instance \Monad\,\Gen\,\where\\
  &\ind \return a = \Gen\,(\lam k.\,k\,a)\\
  &\ind \mit{ga} \bind f = \Gen\,(\lam k.\,\unGen\,\mit{ga}\,(\lam a.\,(f\,a\,k)))
\end{alignat*}
From the $\Monad$ instance, the $\Functor$ and $\Applicative$ instances can be also derived.
We will reuse $(\fmap)$ and $(\ap)$ for applicative notation
as well.
An inhabitant of $\Gen\,A$ can be viewed as an action whose effect is to produce
some object code and returns an $A$ value that can depend on produced object
binders. $\Gen$ can be only ``run'' when the result type is an object type:
\begin{alignat*}{3}
  &\runGen : \Gen\,(\Up A) \to \Up A\\
  &\runGen\,\vma = \unGen\,\vma\,\id
\end{alignat*}
We can let-bind object expressions in $\Gen$:
\begin{alignat*}{3}
  & \gen : \Up A \to \Gen\,(\Up A) \\
  & \gen\,a = \Gen\,(\lam k.\,\ql \letdef x : A := \spl a; \spl(k\,\ql x \qr) \qr)
\end{alignat*}
And also recursive function definitions:
\begin{alignat*}{3}
  & \genRec : \{A : \CTy\} \to (\Up A \to \Up A) \to \Gen\,(\Up A) \\
  & \genRec\,f = \Gen\,(\lam k.\,\qt{\letrec x : A := \spl(f\,\qt{x}); \spl(k\,\qt{x})})
\end{alignat*}
Now, using do-notation, we may write $\mdo \{x \fro \gen\,\qt{10 + 10}; y \fro
\gen\,\qt{20 + 20}\};\return \qt{x + y}\}$, for a $\Gen\,(\Up \Nat)$
action. Running this with $\runGen$ yields $\qt{\letdef x := 10 + 10; \letdef y
  := 20 + 20; x + y}$. We can also define a ``safer'' binding-time improvement for
products, using let-insertion:
\begin{alignat*}{3}
  &\hspace{-4em}\mathrlap{\instance \Improve\,(A,\,B)\,(\Gen\,(\Up A,\,\Up B))\,\where}\\
  &\hspace{-4em}\ind \mup\,x   && = \mdo\{x \fro \gen\,x; \return (\qt{\fst\,\spl x},\,\qt{\snd\,\spl x})\}\\
  &\hspace{-4em}\ind \mdown\,x && = \runGen\,\$\,\mdo \{(a,\,b) \fro x; \return \qt{(\spl a,\, \spl b)}\}
\end{alignat*}

Working in $\Gen$ is convenient, since we can freely generate object code and
also have access to the full metalanguage. Also, the whole point of staging is
that \emph{eventually} all metaprograms will be used for the purpose of code
generation, so ultimately we always want to $\runGen$ our actions. So why not
just always work in $\Gen$? We may not want to do that, because we would like to
reason about the size and makeup of the code we are generating, and the
implicit code generation in $\Gen$ is not very transparent. This is a bit
similar to the $\msf{IO}$ monad in Haskell, where eventually everything needs
to run in $\msf{IO}$, but we may prefer to not write most of our program in
$\msf{IO}$.

@@

Let us start with the $\Maybe$ monad. We have $\data \Maybe\,A :=
\Nothing\,|\,\Just\,A$, and $\Maybe$ itself is available as a $\VTy \to \VTy$
metafunction. However, we cannot directly fashion a monad out of $\Maybe$, since
we do not have enough type formers in $\VTy$. We could try to use the following
type for binding:
\[ (\bind) : \Up(\Maybe\,A) \to (\Up\,A \to \Up\,(\Maybe\,B)) \to \Up(\Maybe\,B) \]
This works, but the definition necessarily uses runtime case splits on $\Maybe$
values, many of which could be optimized away during staging. Also, not having
a ``real'' monad is inconvenient for the purpose of code reuse.

Instead, our strategy is to only use proper monads in $\MTy$, and convert
between object types and meta-monads when necessary, as a form of binding-time
improvement. Assume that $\Maybe_M$ is a meta-level monad, and $\MaybeT_\M$ is
the monad transformer with the standard definition:
\[ \newtype \MaybeT_\M\,M\,A = \MaybeT_\M\,\{\runMaybeT_\M : M\,(\Maybe_\M\,A)\} \]
Then the object-level $\Maybe$ is binding-time improved as follows:
\begin{alignat*}{3}
  &\instance \Improve\,(\Maybe\,A)\,(\MaybeT_\M\,\Gen\,(\Up\,A))\,\where\\
  %% &\ind \mup : \Up\,(\Maybe\,A) \to \MaybeT_\M\,\Gen\,(\Up\,A) \\
  &\ind \mup\,x = \MaybeT_\M \dlr \Gen \dlr \lam k.\\
  &\ind\ind \qt{\case \spl x\,\of \Nothing \to \spl(k\,\Nothing_\M);\Just\,a \to \spl(k\,(\Just_\M\qt{a}))}\\
  %% &\ind \mdown : \MaybeT_\M\,\Gen\,(\Up\,A) \to \Up\,(\Maybe\,A) \\
  &\ind \mdown\,(\MaybeT_\M\,(\Gen\,\vma)) = \\
  &\ind\ind \vma\,(\lam x.\,\case x\,\of \Nothing_\M \to \qt{\Nothing}; \Just_\M\,a \to \qt{\Just\,\spl a})
\end{alignat*}
With this, we get the $\Monad$ instance for free from $\MaybeT_\M$ and $\Gen$. A small example:
\[ \letdef n : \Maybe\,\Nat := \spl(\mdown\,\$\,\mdo \{x \fro \return\,\qt{10}; y \fro \return\,\qt{20}; \return\,\qt{x + y})\});\,... \]
Since $\MaybeT_\M$ is meta-level, its binding always fully computes at staging time. Thus, the above code is staged to
\[ \letdef n : \Maybe\,\Nat := \Just\,(10 + 20);\;... \]
Assume also a $\lift : \Monad\,M \RA M\,A \to \MaybeT_\M\,M\,A$ operation, as in
Haskell. We shall assume this for all monad transformers henceforth. We can do
let-insertion in $\MaybeT_\M\,\Gen$ by simply lifting:
\begin{alignat*}{3}
  &\gen' : \Up A \to \MaybeT_\M\,\Gen\,(\Up A) \\
  &\gen'\,a = \lift\,(\gen\,a)
\end{alignat*}
However, it is more convenient to proceed in the style of Haskell's monad
transformer library \cite{TODO}, and have a class for monads that can do code
generation:
\begin{alignat*}{3}
&\tyclass\,\Monad\,M \RA \MonadGen\,M\,\where\\
&\ind \liftGen : \Gen\,A \to \M\,A\\
&\instance \MonadGen\,\Gen\,\where\,\liftGen = \id \\
&\instance \MonadGen\,M \RA \MonadGen\,(\MaybeT_\M\,M)\,\where\,\liftGen = \lift \circ \liftGen
\end{alignat*}
We also redefine $\gen$ and $\genRec$ to work in any $\MonadGen$, so from now on we have:
\begin{alignat*}{3}
 & \gen   &&: \MonadGen\,M \RA \Up A \to M\,(\Up A)\\
 &\genRec &&: \MonadGen\,M \RA (\Up A \to \Up A) \to M\,(\Up A)
\end{alignat*}
